---
title: "Getting Started with CrossingTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CrossingTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(CrossingTools)

# ------------------------------------------------------------
# Basic Example Setup
# ------------------------------------------------------------
# In this section we create simple simulated data that will be used
# throughout the vignette to illustrate the functionality of CrossingTools.

n_ind   <- 20   # number of individuals
n_mark  <- 200  # number of markers
n_trait <- 3    # number of traits

# Generate a diploid genotype matrix (0 = AA, 2 = BB)
M <- matrix(sample(c(0, 2), n_ind * n_mark, replace = TRUE), nrow = n_ind)

# Simulate additive marker effects for all traits (markers x traits)
U <- matrix(rnorm(n_mark * n_trait), ncol = n_trait)

# Calculate individual breeding values (genetic values)
A <- M %*% U
```

## Desired Gains Index (DGI)

CrossingTools supports multi-trait optimization via a Desired Gains Index (DGI).
The DGI translates *target improvements* (desired gains) into *trait weights*,
which can then be used to compute an index for individuals or crosses.

In this example we aim to improve all traits equally.

```{r desired-gains}
# Desired gains: improve all traits by the same amount
gains <- rep(10, n_trait)

# Calculate desired gains index for individuals and the resulting trait weights
DG <- calculate_desired_gains(
  A            = A,           # breeding values per individual x trait
  use.V.approx = TRUE,        # uses cov(A) as an empirical approximation
  gains        = gains
)

# Index score per individual
head(DG$index)

# Trait weights implied by desired gains
DG$weight
```

```{r genetic map and crosses}
# Define a simple two-chromosome map with evenly spaced markers in Morgan
map <- data.frame(
  site = 1:n_mark,
  chr  = rep(1:2, each = n_mark / 2),
  pos  = c(seq(0, 1.5, length.out = n_mark / 2),
           seq(0, 1.5, length.out = n_mark / 2))
)

# Generate all possible pairwise crosses among the individuals.
# This scales as O(n^2) and can become large in realistic populations.
crosses <- t(combn(n_ind, 2))
```

### Doubled Haploids (DH)

The Doubled Haploid (DH) population type assumes complete homozygosity.
The function `calculate_variances_DH()` estimates expected genetic variance,
superior progeny value (SPV), and (optionally) trait covariances for all possible crosses.

With the updated API, you can additionally request an **index** at the cross level
by setting `calculate.index = TRUE` and providing `weights` (e.g., from the DGI).

```{r dh}
out_dh <- calculate_variances_DH(
  crosses         = crosses,
  genetic.map     = map,        # must contain site, chr, pos covering all M columns
  M               = M,          # genotype matrix (0/2 coding)
  U               = U,          # additive marker effects (markers x traits)
  t               = 2,          # random-mating generations before DH creation
  intensity       = 1,          # selection differential for SPV
  covariance      = TRUE,       # include segregation covariance matrices
  calculate.index = TRUE,       # compute index statistics per cross
  weights         = DG$weight,  # trait weights for index
  method          = "osthushenrich"
)

# The cross_values component summarizes per-cross statistics per trait:
#  * EGBV: Expected Genetic Breeding Value per trait
#  * var:  Genetic variance of segregating progeny per trait
#  * SPV:  Superior Progeny Value per trait
head(out_dh$cross_values)

# If covariance = TRUE, each cross includes a trait covariance matrix.
# The following example shows the first cross (trait x trait covariance):
out_dh$covariances[[1]]

# The index component summarizes the same quantities for the multi-trait index:
#  * EGBV: expected index mean
#  * var:  index variance
#  * SPV:  index-based superior progeny value
head(out_dh$index)
```

### Recombinant Inbred Lines (RIL)

The function `calculate_variances_RIL()` works analogously to the DH version,
but models the genetic variance expected after RIL generation instead of DH.

```{r ril}
out_ril <- calculate_variances_RIL(
  crosses         = crosses,
  genetic.map     = map,
  M               = M,
  U               = U,
  t               = 2,
  intensity       = 1,
  covariance      = TRUE,
  calculate.index = TRUE,
  weights         = DG$weight,
  method          = "osthushenrich"
)

# Summarize the results
head(out_ril$cross_values)

# Inspect the first covariance matrix (trait x trait)
out_ril$covariances[[1]]

# Summarize the index
head(out_ril$index)
```

### Four-Way Crosses (DH & RIL)

Four-way (and three-way) crosses are constructed by combining multiple biparental crosses.
Columns 1:2 correspond to the first pair, and columns 3:4 to the second pair.
For a 3-way cross, the first pair can be duplicated in columns 1:2, with new parents
in columns 3 and/or 4.

The four-way variants work similarly and can also compute an index.

```{r four-way}
crosses4 <- t(combn(n_ind, 4))

out_4w_DH <- calculate_variances_4W_DH(
  crosses         = crosses4,
  genetic.map     = map,
  M               = M,
  U               = U,
  t               = 1,
  intensity       = 1,
  covariance      = TRUE,
  calculate.index = TRUE,
  weights         = DG$weight
)

out_4w_RIL <- calculate_variances_4W_RIL(
  crosses         = crosses4,
  genetic.map     = map,
  M               = M,
  U               = U,
  t               = 1,
  intensity       = 1,
  covariance      = TRUE,
  calculate.index = TRUE,
  weights         = DG$weight
)

head(out_4w_DH$cross_values)
head(out_4w_RIL$cross_values)
```

### Clonal Breeding with Heterozygous Parents and F1s (Additive + Dominance)

For clonal breeding with heterozygous parents, both additive and dominance effects
contribute to offspring performance. To model this, provide haplotype matrices for
each parent: `hap1` and `hap2` (0/1 coding), as well as additive (`U`) and dominance (`D`)
marker effects.

```{r f1}
hap1 <- matrix(sample(0:1, n_ind * n_mark, replace = TRUE), nrow = n_ind, ncol = n_mark)
hap2 <- matrix(sample(0:1, n_ind * n_mark, replace = TRUE), nrow = n_ind, ncol = n_mark)

# Simulate dominance marker effects (markers x traits)
D <- matrix(rnorm(n_mark * n_trait), ncol = n_trait)

out_f1 <- calculate_variances_F1(
  crosses         = crosses,
  genetic.map     = map,
  hap1            = hap1,
  hap2            = hap2,
  U               = U,
  D               = D,
  intensity       = 1,
  covariance      = TRUE,
  calculate.index = TRUE,
  weights         = DG$weight
)

# The results include additive and dominance contributions to genetic variance
# and (if covariance = TRUE) additive/dominance covariance matrices per cross.
head(out_f1$cross_values)

# Example: first cross additive and dominance covariance matrices (trait x trait)
out_f1$additive_covariances[[1]]
out_f1$dominance_covariances[[1]]

# Index summary per cross
head(out_f1$index)
```

### Optimal Cross Selection (OCS)

OCS identifies an optimal subset of crosses balancing expected genetic gain and diversity
using a genetic algorithm (GA). It can use any cross-level criterion computed above
(e.g., EGBV, SPV, or the index SPV).

```{r ocs}
# Compute a genomic relationship matrix (G) as a measure of relatedness.
G <- tcrossprod(scale(M, center = TRUE, scale = FALSE)) / n_mark

# Select a cross-level criterion (example: SPV for trait 1).
crit <- out_dh$cross_values$SPV1

ocs <- optimal_cross_selection(
  crosses        = crosses,
  ncrosses       = 20,
  G              = G,
  u              = crit,
  target.angle   = 30 * pi / 180,  # trade-off between gain and diversity (radians)
  n.select       = 200,            # GA parent pool size
  n.pop          = 1000,           # GA population size
  max.generation = 200,            # max GA iterations
  max.iteration  = 50,             # early stopping threshold
  angle.penalty  = 0.01,           # penalty for angle deviation
  n.Threads      = 2L
)

# The selected optimal cross plan (ncrosses x 3: parent 1; parent 2; the row in crosses which it refers to) is returned as:
head(ocs$crossPlan)
```

### Other Utilities

#### Optimal Haploid Value (OHV)

The OHV function identifies optimal haploid gamete combinations given additive marker effects.
It can operate using defined marker blocks or treat each marker independently if none are provided.

```{r ohv}
ohv <- calculate_optimal_haploid_value(
  crosses = crosses,
  M       = M,
  U       = U
)

head(ohv)
```

#### Backsolving Marker Effects

If you have individual-level genetic values (e.g., BLUPs or GEBVs),
`u_from_g()` can estimate underlying marker effects `U`. This is useful
for reconstructing marker effects consistent with a known relationship matrix.

```{r backsolve}
u_fit <- u_from_g(
  M              = M,
  G              = G,
  g              = A,
  scaling.factor = 1 / n_mark # scaling factor applied to the GRM
)

# `u_fit` contains estimated marker effects and (depending on settings) variance components.
head(u_fit)
```
